---
title: "Introduction to postcard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to postcard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Setup-chunk to load the package, set a seed and turn off verbosity for the rendering of the vignette.

```{r setup, message = FALSE}
library(postcard)
withr::local_seed(1395878)
withr::local_options(list(postcard.verbose = 0))
```

postcard provides tools for accurately estimating marginal effects using plug-in estimation with GLMs, including increasing precision using prognostic covariate adjustment. See [Powering RCTs for marginal effects with GLMs using prognostic score adjustment](https://arxiv.org/abs/2503.22284) by Højbjerre-Frandsen et. al (2025).

# Plug-in estimation of marginal effects and variance estimation using influence functions

The use of plug-in estimation and influence functions can help us obtain more accurate estimates. Coupled with prognostic covariate adjustment, we can increase the precision of our estimates and obtain a higher power with sacrificing control over the type I error rate.

Introductory examples on the use of `rctglm()` and `rctglm_with_prognosticscore()` functions are available here. For more details, see `vignette("model-fit")`.

## Simulating data for exploratory analyses

First, we simulate some data to be able to enable showcasing of the functionalities. For this we use the `glm_data()` function from the package, where the user can specify an expression alongside variables and a family of the response to then simulate a response from a GLM with linear predictor given by the expression provided.

```{r dat-sim}
n <- 1000
b0 <- 1
b1 <- 3
b2 <- 2

# Simulate data with a non-linear effect
dat_treat <- glm_data(
  Y ~ b0+b1*sin(W)^2+b2*A,
  W = runif(n, min = -2, max = 2),
  A = rbinom(n, 1, prob = 1/2),
  family = gaussian() # Default value
)
```

## Fitting `rctglm()` without prognostic covariate adjustment

The `rctglm()` function estimates any specified estimand using plug-in estimation for randomised clinical trials and estimates the variance using the influence function of the marginal effect estimand. 

The interface of `rctglm()` is similar to that of the `stats::glm()` function but with an added mandatory specification of

-   `exposure_indicator`: The randomisation variable in data, usually being the (name of) the treatment variable
-   `exposure_prob`: The randomisation ratio - the probability of being allocated to group 1 (rather than 0)
    -   As a default, a ratio of 1's in data is used
-   `estimand_fun`: An estimand function
    -   As a default, the function takes the average treatment effect (ATE) as the estimand

Thus, we can estimate the ATE by simply writing the below:

> Note that as a default, `verbose = 2`, meaning that information about the algorithm is printed to the console. However, here we suppress this behavior. See more in `vignette("model-fit")`.

```{r noprog-run}
ate <- rctglm(formula = Y ~ A * W,
              exposure_indicator = A,
              exposure_prob = 1/2,
              data = dat_treat,
              family = "gaussian") # Default value
```

This creates an `rctglm` object which prints as
```{r noprog-show}
ate
```

The structure of such an `rctglm` object is broken down in the **`Value`** section of the documentation in `rctglm()`.
    
Methods available are `estimand` (or the shorthand `est`) which prints a `data.frame` with and estimate of the estimand and its standard error. A method for the generics `coef` and `predict` are also available that are "shortcuts" to applying the corresponding methods to the underlying `glm` fit.
```{r}
est(ate)
```

See more info in the documentation page `rctglm_methods()`.

## Using prognostic covariate adjustment

The `rctglm_with_prognosticscore()` function uses the `fit_best_learner()` function to fit a prognostic model to historical data and then uses the prognostic model to predict
\begin{align}
\mathbb{E}[Y|X,A=0]
\end{align}

for all observations in the current data set. These *prognostic scores* are then used as a covariate in the GLM when running `rctglm()`.

Allowing the use of complex non-linear models to create such a prognostic score allows utilising information from potentially many variables, "catching" non-linear relationships and then using all this information in the GLM model using a single covariate adjustment.

We simulate some historical data to showcase the use of this function as well:
```{r hist-data}
dat_notreat <- glm_data(
  Y ~ b0+b1*sin(W)^2,
  W = runif(n, min = -2, max = 2),
  family = gaussian # Default value
)
```

The call to `rctglm_with_prognosticscore()` is the same as to `rctglm()` but with an added specification of

-   (Historical) data to fit the prognostic model using `fit_best_learner()`
-   A formula used when fitting the prognostic model
    -   Default uses all covariates in the data.
-   (Optionally) number folds in cross validation and a list of learners for fitting the best learner

Thus, a simple call which estimates the average treatment effect, adjusting for a prognostic score, is seen below:
```{r prog-run}
ate_prog <- rctglm_with_prognosticscore(
  formula = Y ~ A * W,
  exposure_indicator = A,
  exposure_prob = 1/2,
  data = dat_treat,
  family = gaussian(link = "identity"), # Default value
  data_hist = dat_notreat)
```

Quick results of the fit can be seen by printing the object:
```{r prog-show}
ate_prog
```

It's evident that in this case where there is a non-linear relationship between the covariate we observe and the response, adjusting for the prognostic score reduces the standard error of our estimand approximation by quite a bit.

### Investigating the prognostic model

Information on the prognostic model is available in the list element `prognostic_info`, which the method `prog()` can be used to extract. A breakdown of what this list includes, see the **`Value`** section of the `rctglm_with_prognosticscore()` documentation.

# Power approximation

In cases of seeking to conduct new studies, sample size/power analyses are vital to the successful planning of such studies. Here, we present implementations in this package that take advantage of power approximation formulas to perform such analyses.

See a more detailed walkthrough of a use case in `vignette("prospective-power")`.

## For marginal effects

The method proposed in [Powering RCTs for marginal effects with GLMs using prognostic score adjustment](https://arxiv.org/abs/2503.22284) by Højbjerre-Frandsen et. al (2025), which can be used to estimate the power when estimating any marginal effect, is implemented in the function `power_marginaleffect()`.

According to the conservative approach in the article, if wanting to conduct power analyses to figure out how many participants is needed for an upcoming trial, where you are planning to use *prognostic covariate adjustment*, predictions should be obtained from a discrete super learner identical to the one planned to use for generating prognostic scores when adjusting in the analysis when estimating the marginal effect.

> Here we showcase a simple use of a `glm()`, but fx. `fit_best_learner()` could be used to fit a discrete super learner as the prediction model. Could also add steps to get out-of-sample (OOS) predictions (see examples).

```{r}
pred_mod <- glm(Y ~ W + A, data = dat_treat)
preds <- predict(pred_mod, dat_treat)

power_marginaleffect(
  response = dat_treat$Y,
  predictions = preds,
  target_effect = 0.4,
  exposure_prob = 1/2
)
```

## Specific to linear models

### Estimating the assumed variance

Finding the assumed variance to use for your power analysis with an ANCOVA model can be done using the `variance_ancova` function, which estimates the term $\sigma^2(1-R^2)$ given a `formula` and `data`.

```{r}
vanc <- variance_ancova(Y ~ A + W, data = dat_treat)
vanc
```

### Finding the power (or sample size)

Functions `power_gs()` and `power_nc()` exist, which estimate the power given a sample size `n` using approximation formulas. The functions are the results of two different approximation formulas but behave exactly the same except for a mandatory specification of a `df` argument for the `power_nc` function, which gives the degrees of freedom in the t-distribution used.

> Details about the formulas are available in the documentation

For the Guenther-Schouten approximation, the formula directly gives us a sample size as a function of the power, so getting the required sample size as a function of the power is available in the function `samplesize_gs()`.

```{r}
power_gs(variance = vanc, n = 100, ate = 0.8)
power_nc(variance = vanc, n = 100, df = 97, ate = 0.8)
samplesize_gs(variance = vanc, ate = 0.8, power = 0.9)
```
