---
title: "Prospective Power Estimation"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Prospective Power Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r setup, message = FALSE}
library(postcard)
withr::local_seed(1395878)
withr::local_options(list(postcard.verbose = 0))
```

# Estimating the power for marginal effects

The method proposed in [Powering RCTs for marginal effects with GLMs using prognostic score adjustment](https://arxiv.org/abs/2503.22284) by Højbjerre-Frandsen et. al (2025), which can be used to estimate the power when estimating any marginal effect, is implemented in the function `power_marginaleffect()`.

An introductory example is available in `vignette("postcard")`, but here we describe how to specify arguments to change the default behavior of the function to align with assumptions for the power estimation.

### Simulating some data

We generate count data to showcase the flexibility of `power_marginaleffect()` that does not assume a linear model.

```{r}
n_train <- 2000
n_test <- 200
b1 <- 1.2
b2 <- 2.3
b3 <- 1
b4 <- 1.8

train_pois <- dplyr::mutate(
  glm_data(
    Y ~ b1*abs(sin(X1))-b2*X2+b3*X3-b4*X2*X3,
    X1 = runif(n_train, min = -2, max = 2),
    X2 = rnorm(n_train, mean = 2, sd = 2),
    X3 = rgamma(n_train, shape = 1),
    family = gaussian()
  ),
  Y = round(abs(Y))
)
test_pois <- dplyr::mutate(
  glm_data(
    Y ~ b1*abs(sin(X1))-b2*X2+b3*X3-b4*X2*X3,
    X1 = runif(n_test, min = -2, max = 2),
    X2 = rnorm(n_test, mean = 2, sd = 2),
    X3 = rgamma(n_test, shape = 1),
    family = gaussian()
  ),
  Y = round(abs(Y))
)
```

## Controlling assumptions

As a default, the variance in group 0 is estimated from the sample variance of the response, and the variance in group 1 is assumed to be the same as the estimated variance in group 0. Use argument `var1` to change the variance estimate in group 1, either with a function that modifies the estimate obtained for group 0 or as a `numeric`.

The same is true for the MSE, where `kappa1_squared` as a default is taken to be the same as the MSE in group 0 unless a `function` or `numeric` is specified.

Below is an example showcasing the specification of `var1` and `kappa1_squared` according to prior beliefs.

> Note that we fit a prognostic model using `fit_best_learner()` and use this as our prediction in accordance with the guidance in the main reference to obtain a conservative power estimate of estimating a marginal effect in a GLM that adjusts for the prognostic score as a covariate alongside other covariates.

```{r}
lrnr <- fit_best_learner(list(mod = Y ~ X1 + X2 + X3), data = train_pois)
preds <- dplyr::pull(predict(lrnr, new_data = test_pois))

power_marginaleffect(
  response = test_pois$Y,
  predictions = preds,
  var1 = function(var0) 1.2 * var0,
  kappa1_squared = 2,
  estimand_fun = "rate_ratio",
  target_effect = 1.3,
  exposure_prob = 1/2
)
```

The function `repeat_power_marginaleffect()` allows passing arguments along to `power_marginaleffect()`, so we are able to quickly create power curves for the above case.

We define our models in a named list and define a function of a single parameter - the sample size `n` - that simulates the test data with the same structure as above but for different sample sizes. 

```{r}
ancova_prog_list <- list(
  ANCOVA = glm(Y ~ X1 + X2 + X3, data = train_pois, family = poisson),
  # "Null model" = glm(Y ~ 1, data = train_pois, family = poisson),
  "ANCOVA with prognostic score" = fit_best_learner(list(mod = Y ~ X1 + X2 + X3), data = train_pois)
)
test_pois_fun <- function(n) {
  dplyr::mutate(
    glm_data(
      Y ~ b1*abs(sin(X1))-b2*X2+b3*X3-b4*X2*X3,
      X1 = runif(n, min = -2, max = 2),
      X2 = rnorm(n, mean = 2, sd = 2),
      X3 = rgamma(n, shape = 1),
      family = gaussian()
    ),
    Y = round(abs(Y))
  )
}
```

We then run the power estimation and plot the results:

```{r}
rpm <- repeat_power_marginaleffect(
  model_list = ancova_prog_list,
  test_data_fun = test_pois_fun,
  ns = 10:300, n_iter = 1,
  var1 = function(var0) 1.2 * var0,
  kappa1_squared = 2,
  estimand_fun = "rate_ratio",
  target_effect = 1.3,
  exposure_prob = 1/2
)

plot(rpm)
```



